---
title: "Advanced Topics: Multilevel, Measurement Error, and Gaussian Processes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced Topics: Multilevel, Measurement Error, and Gaussian Processes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette covers advanced modeling techniques:

- **Multilevel models**: Varying intercepts, slopes, and complex structures
- **Measurement error**: Handling imperfect measurements
- **Gaussian processes**: Flexible non-parametric regression
- **Networks**: Social network analysis

```{r setup}
library(tidyrethinking)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Multilevel Models

### Varying Intercepts

Groups differ in their baseline levels:

```{r mlm-intercepts}
# Simulate grouped data
ml_data <- simulate_varying_intercepts(
  n_groups = 20,
  n_per_group = 10,
  grand_intercept = 5,
  between_group_sd = 2,
  within_group_sd = 1,
  seed = 42
)

# Visualize
ggplot(ml_data, aes(x = factor(group), y = y)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "red") +
  labs(
    x = "Group",
    y = "Outcome",
    title = "Varying Intercepts Across Groups",
    subtitle = "Red line = grand mean"
  ) +
  theme_rethinking() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Partial Pooling (Shrinkage)

Multilevel models balance between complete pooling and no pooling:

```{r mlm-shrinkage}
# Compute group statistics
group_stats <- ml_data %>%
  group_by(group) %>%
  summarise(
    mean_y = mean(y),
    n = n()
  )

# Compute shrinkage
shrinkage <- compute_shrinkage(
  group_means = group_stats$mean_y,
  group_ns = group_stats$n,
  within_sd = 1,
  between_sd = 2,
  grand_mean = 5
)

# Visualize
ggplot(shrinkage, aes(x = observed_mean, y = shrunken_mean)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_segment(aes(xend = shrunken_mean, yend = shrunken_mean),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "steelblue", alpha = 0.6) +
  geom_point(aes(size = n, color = shrinkage_factor), alpha = 0.8) +
  scale_color_viridis_c(option = "plasma") +
  labs(
    x = "Observed group mean (no pooling)",
    y = "Shrunken estimate (partial pooling)",
    color = "Shrinkage",
    size = "Group size",
    title = "Partial Pooling Shrinks Estimates Toward Grand Mean",
    subtitle = "Smaller groups shrink more"
  ) +
  theme_rethinking()
```

### Intraclass Correlation (ICC)

How much variance is between vs within groups?

```{r mlm-icc}
icc_result <- compute_icc(
  between_group_sd = 2,
  within_group_sd = 1
)

icc_result

# ICC interpretation:
# 0.0: All variance within groups (no clustering)
# 1.0: All variance between groups (perfect clustering)
# 0.5: Equal variance between and within
```

### Varying Slopes

Groups differ in their response to predictors:

```{r mlm-slopes}
# Simulate varying slopes
slopes_data <- simulate_varying_slopes(
  n_groups = 15,
  n_per_group = 20,
  grand_intercept = 3,
  grand_slope = 0.5,
  slope_sd = 0.3,
  residual_sd = 0.5,
  seed = 42
)

# Visualize
ggplot(slopes_data, aes(x = x, y = y, group = group, color = group)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  scale_color_viridis_c() +
  labs(
    x = "Predictor",
    y = "Outcome",
    title = "Varying Slopes Across Groups",
    subtitle = "Each group has different relationship"
  ) +
  theme_rethinking() +
  theme(legend.position = "none")
```

### Cross-Classified Models

Non-nested groupings (e.g., students in schools AND neighborhoods):

```{r mlm-cross}
# Simulate cross-classified structure
cross_data <- simulate_cross_classified(
  n_groups_1 = 10,  # Schools
  n_groups_2 = 8,   # Neighborhoods
  n_per_combo = 5,
  group_1_sd = 1,
  group_2_sd = 0.8,
  seed = 42
)

# Visualize structure
ggplot(cross_data, aes(x = factor(group_1), y = factor(group_2))) +
  geom_count(aes(size = ..n..), color = "steelblue", alpha = 0.6) +
  scale_size_area(max_size = 10) +
  labs(
    x = "Group 1 (e.g., School)",
    y = "Group 2 (e.g., Neighborhood)",
    size = "N observations",
    title = "Cross-Classified Structure",
    subtitle = "Each cell represents a school × neighborhood combination"
  ) +
  theme_rethinking()
```

### Temporal Autocorrelation

Observations within groups are correlated over time:

```{r mlm-temporal}
# Simulate AR(1) within groups
temporal_data <- simulate_temporal_autocorrelation(
  n_groups = 5,
  n_timepoints = 50,
  rho = 0.7,  # AR(1) coefficient
  sigma = 1,
  seed = 42
)

# Visualize time series
ggplot(temporal_data, aes(x = time, y = y, color = factor(group))) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1, alpha = 0.5) +
  scale_color_viridis_d() +
  labs(
    x = "Time",
    y = "Outcome",
    color = "Group",
    title = "Temporal Autocorrelation Within Groups",
    subtitle = sprintf("AR(1) coefficient ρ = %.1f", 0.7)
  ) +
  theme_rethinking()

# Check autocorrelation
temporal_data %>%
  group_by(group) %>%
  summarise(
    acf_lag1 = cor(y[-n()], y[-1]),
    .groups = "drop"
  )
```

## Measurement Error

### Understanding Attenuation Bias

Measurement error in predictors biases estimates toward zero:

```{r measurement-error}
# Simulate measurement error
error_data <- simulate_measurement_error(
  n = 500,
  true_effect = 0.8,
  measurement_sd = 0.5,
  seed = 42
)

# Naive regression (ignores measurement error)
naive_fit <- lm(y ~ x_observed, data = error_data)
naive_estimate <- coef(naive_fit)["x_observed"]

# True relationship (if we could observe x perfectly)
true_fit <- lm(y ~ x_true, data = error_data)
true_estimate <- coef(true_fit)["x_true"]

# Expected bias
bias <- compute_attenuation_bias(
  true_effect = 0.8,
  true_sd = 1,
  measurement_sd = 0.5
)

tibble(
  Model = c("True", "Observed", "Expected bias"),
  Estimate = c(true_estimate, naive_estimate, bias$attenuated_effect)
) %>%
  knitr::kable(digits = 3)
```

### Visualizing Measurement Error

```{r measurement-viz}
ggplot(error_data, aes(x = x_true, y = y)) +
  geom_point(aes(color = "True X"), alpha = 0.3, size = 2) +
  geom_smooth(aes(color = "True X"), method = "lm", se = FALSE, linewidth = 1.5) +
  geom_point(aes(x = x_observed, color = "Observed X"), alpha = 0.3, size = 2) +
  geom_smooth(aes(x = x_observed, color = "Observed X"), 
              method = "lm", se = FALSE, linewidth = 1.5) +
  scale_color_manual(values = c("True X" = "darkgreen", "Observed X" = "coral")) +
  labs(
    x = "Predictor",
    y = "Outcome",
    color = "Data",
    title = "Measurement Error Attenuates Estimates",
    subtitle = sprintf("True slope = %.2f, Observed slope = %.2f", 
                      true_estimate, naive_estimate)
  ) +
  theme_rethinking()
```

## Missing Data

### Missing Completely at Random (MCAR)

Missingness unrelated to any variable:

```{r missing-mcar}
# Simulate MCAR
mcar_data <- simulate_mcar(
  n = 500,
  true_effect = 0.5,
  missing_prob = 0.3,
  seed = 42
)

# Complete case analysis is unbiased
complete_cases <- mcar_data %>% filter(!missing)
lm(y ~ x, data = complete_cases) %>% coef()
```

### Missing at Random (MAR)

Missingness depends on observed variables:

```{r missing-mar}
# Simulate MAR
mar_data <- simulate_mar(
  n = 500,
  true_effect = 0.5,
  covariate_effect_on_missing = 1.5,
  seed = 42
)

# Missingness depends on covariate
table(mar_data$missing, mar_data$z > 0)

# Complete case analysis may be biased
# Need to model missingness or use imputation
```

### Instrumental Variables

Use an instrument to handle confounding or measurement error:

```{r missing-iv}
# Simulate instrumental variable design
iv_data <- simulate_instrument(
  n = 1000,
  instrument_strength = 0.7,  # Z -> X
  true_effect = 0.5,           # X -> Y
  confounder_effect = 1.0,     # U -> X, U -> Y
  seed = 42
)

# Naive regression (biased by confounding)
naive_iv <- lm(y ~ x, data = iv_data)
coef(naive_iv)["x"]

# Two-stage least squares (unbiased)
tsls_result <- compute_tsls(
  data = iv_data,
  y_var = "y",
  x_var = "x",
  z_var = "z"
)

tsls_result
```

## Gaussian Processes

### Squared Exponential Kernel

Smooth, continuous functions:

```{r gp-kernel}
# Define observation points
x_obs <- c(-2, -1, 0, 1.5, 2.5)
y_obs <- c(0, 1, 0.5, -0.5, 0)

# Compute covariance
cov_matrix <- compute_squared_exponential(
  x1 = x_obs,
  x2 = x_obs,
  eta_sq = 1,    # Maximum covariance
  rho_sq = 1     # Length scale
)

# Visualize covariance
cov_df <- as.data.frame(cov_matrix)
colnames(cov_df) <- paste0("x", seq_along(x_obs))
cov_df$x1 <- x_obs

cov_long <- cov_df %>%
  pivot_longer(-x1, names_to = "x2", values_to = "covariance") %>%
  mutate(x2 = as.numeric(sub("x", "", x2)),
         x2_val = x_obs[x2])

ggplot(cov_long, aes(x = x1, y = x2_val, fill = covariance)) +
  geom_tile() +
  scale_fill_gradient2(low = "white", high = "steelblue", mid = "lightblue", midpoint = 0.5) +
  labs(
    x = "Point i",
    y = "Point j",
    fill = "Covariance",
    title = "Squared Exponential Kernel",
    subtitle = "Closer points are more correlated"
  ) +
  theme_rethinking()
```

### Prior Samples from GP

```{r gp-prior}
# Sample functions from GP prior
x_grid <- seq(-3, 3, length.out = 100)
prior_samples <- sample_gp_prior(
  x = x_grid,
  n_samples = 20,
  eta_sq = 1,
  rho_sq = 1,
  seed = 42
)

# Visualize
ggplot(prior_samples, aes(x = x, y = y, group = sample)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  labs(
    x = "Input",
    y = "Output",
    title = "Prior Functions from Gaussian Process",
    subtitle = "Each line is one possible function"
  ) +
  theme_rethinking()
```

### GP Regression (Posterior)

```{r gp-posterior}
# Compute posterior predictions
x_pred <- seq(-3, 3, length.out = 100)
posterior <- compute_gp_posterior(
  x_obs = x_obs,
  y_obs = y_obs,
  x_pred = x_pred,
  eta_sq = 1,
  rho_sq = 1,
  sigma_sq = 0.1
)

# Visualize
ggplot(posterior, aes(x = x)) +
  geom_ribbon(aes(ymin = mean - 2*sd, ymax = mean + 2*sd),
              fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = mean), color = "steelblue", linewidth = 1) +
  geom_point(data = tibble(x = x_obs, y = y_obs),
             aes(x = x, y = y), size = 3, color = "red") +
  labs(
    x = "Input",
    y = "Output",
    title = "Gaussian Process Regression",
    subtitle = "Line = posterior mean, ribbon = 95% credible interval, points = observations"
  ) +
  theme_rethinking()
```

### Periodic Kernel

For cyclic patterns (seasons, days, etc.):

```{r gp-periodic}
# Generate seasonal data
x_seasonal <- seq(0, 3, length.out = 200)  # 3 years
cov_periodic <- compute_periodic_covariance(
  x1 = x_seasonal,
  x2 = x_seasonal,
  eta_sq = 1,
  rho_sq = 0.5,
  period = 1  # Annual cycle
)

# Sample from periodic GP
seasonal_samples <- sample_gp_prior(
  x = x_seasonal,
  n_samples = 10,
  eta_sq = 1,
  rho_sq = 0.5,
  kernel = "periodic",
  period = 1,
  seed = 42
)

ggplot(seasonal_samples, aes(x = x, y = y, group = sample)) +
  geom_line(alpha = 0.5, color = "darkgreen") +
  labs(
    x = "Time (years)",
    y = "Value",
    title = "Periodic Gaussian Process",
    subtitle = "Captures repeating seasonal patterns"
  ) +
  theme_rethinking()
```

## Social Networks

### Dyadic Data

Network data with sender and receiver effects:

```{r network-dyadic}
# Simulate social network
network <- simulate_dyadic_network(
  n_nodes = 20,
  sender_sd = 0.5,
  receiver_sd = 0.5,
  reciprocity = 1.5,
  seed = 42
)

head(network)

# Visualize as matrix
adj_matrix <- create_adjacency_matrix(network, value_col = "tie")

# Convert to long format for heatmap
adj_long <- as.data.frame(adj_matrix) %>%
  mutate(sender = rownames(adj_matrix)) %>%
  pivot_longer(-sender, names_to = "receiver", values_to = "tie")

ggplot(adj_long, aes(x = receiver, y = sender, fill = factor(tie))) +
  geom_tile(color = "white") +
  scale_fill_manual(values = c("0" = "white", "1" = "steelblue")) +
  labs(
    x = "Receiver",
    y = "Sender",
    fill = "Tie",
    title = "Social Network Adjacency Matrix"
  ) +
  theme_rethinking() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6),
        axis.text.y = element_text(size = 6))
```

### Network Statistics

```{r network-stats}
# Compute descriptive statistics
stats <- compute_network_stats(network)
stats

# Interpretation:
# - Density: proportion of possible ties that exist
# - Reciprocity: tendency for ties to be mutual
# - Degree stats: distribution of connections per node
```

### Transitive Closure

"Friend of a friend" becomes a friend:

```{r network-transitive}
# Simulate transitivity
trans_network <- simulate_transitive_closure(
  n_nodes = 15,
  initial_ties = 20,
  transitivity_prob = 0.7,
  seed = 42
)

# Count triangles
triangles <- trans_network %>%
  group_by(sender) %>%
  summarise(n_ties = n(), .groups = "drop")

ggplot(triangles, aes(x = n_ties)) +
  geom_histogram(binwidth = 1, fill = "coral", alpha = 0.7) +
  labs(
    x = "Number of outgoing ties",
    y = "Count of nodes",
    title = "Degree Distribution with Transitivity",
    subtitle = "Some nodes have many ties due to transitivity"
  ) +
  theme_rethinking()
```

## Integration Example

Combining techniques:

```{r integration, eval=FALSE}
# 1. Simulate multilevel data with measurement error
ml_error <- simulate_varying_intercepts(n_groups = 20, n_per_group = 20) %>%
  mutate(
    x_true = rnorm(n()),
    x_obs = x_true + rnorm(n(), 0, 0.5),  # Measurement error
    y = y + 0.7 * x_true + rnorm(n())
  )

# 2. Fit multilevel model with measurement error correction
# (Requires specialized packages like brms or Stan)

# 3. Use GP for non-linear effects
# 4. Account for missing data patterns
# 5. Validate with posterior predictive checks
```

## Resources

- McElreath, R. (2020). *Statistical Rethinking*, Chapters 12-18
- Rasmussen & Williams (2006). *Gaussian Processes for Machine Learning*
- Snijders & Bosker (2012). *Multilevel Analysis*
- Wasserman & Faust (1994). *Social Network Analysis*

## Next Steps

- See `vignette("bayesian-workflow")` for implementing these models with brms
- See `vignette("causal-inference")` for combining DAGs with multilevel models
- Explore **brms** for Bayesian multilevel models
- Explore **tidygraph** and **ggraph** for network visualization
