---
title: "Bayesian Workflow with tidyrethinking and brms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Workflow with tidyrethinking and brms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = FALSE  # Set to TRUE if brms/cmdstan installed
)
```

## Introduction

This vignette demonstrates how to integrate **tidyrethinking** with modern Bayesian modeling packages for a complete workflow:

- **brms**: Full Bayesian regression models using Stan
- **tidybayes**: Extract and visualize posterior draws
- **ggdist**: Enhanced visualizations of distributions
- **bayesplot**: MCMC diagnostics and posterior checks
- **posterior**: Work with posterior distributions

## Setup

```{r setup}
library(tidyrethinking)
library(dplyr)
library(ggplot2)

# Bayesian ecosystem (install separately)
library(brms)
library(tidybayes)
library(ggdist)
library(bayesplot)
library(posterior)
```

## Example: Multilevel Model with brms

### Simulate Data

Use tidyrethinking to generate realistic multilevel data:

```{r simulate}
# Simulate varying intercepts data
set.seed(42)
ml_data <- simulate_varying_intercepts(
  n_groups = 20,
  n_per_group = 10,
  grand_intercept = 5,
  between_group_sd = 2,
  within_group_sd = 1,
  seed = 42
)

head(ml_data)
```

### Fit Model with brms

```{r fit-brms}
# Fit multilevel model
fit <- brm(
  y ~ 1 + (1 | group),
  data = ml_data,
  family = gaussian(),
  prior = c(
    prior(normal(0, 10), class = Intercept),
    prior(exponential(1), class = sd),
    prior(exponential(1), class = sigma)
  ),
  iter = 2000,
  warmup = 1000,
  chains = 4,
  cores = 4,
  seed = 42
)

summary(fit)
```

### Extract Posterior with tidybayes

```{r tidybayes}
# Extract posterior draws in tidy format
posterior_draws <- fit %>%
  spread_draws(b_Intercept, sigma, sd_group__Intercept)

head(posterior_draws)

# Extract group-specific intercepts
group_effects <- fit %>%
  spread_draws(r_group[group, term])

head(group_effects)
```

### Visualize with ggdist

```{r ggdist-viz}
# Plot posterior distributions
posterior_draws %>%
  ggplot(aes(x = sd_group__Intercept)) +
  stat_halfeye(
    .width = c(0.66, 0.95),
    fill = "skyblue"
  ) +
  labs(
    x = "Between-group SD",
    y = "Density",
    title = "Posterior Distribution of Group Variability"
  ) +
  theme_rethinking()

# Compare observed vs fitted group means
group_means <- ml_data %>%
  group_by(group) %>%
  summarise(observed = mean(y))

group_effects %>%
  mutate(fitted = b_Intercept + r_group) %>%
  median_qi(.width = 0.89) %>%
  left_join(group_means, by = "group") %>%
  ggplot(aes(y = group)) +
  geom_interval(aes(xmin = .lower, xmax = .upper)) +
  geom_point(aes(x = fitted), size = 2) +
  geom_point(aes(x = observed), color = "red", shape = 4, size = 3) +
  labs(
    x = "Group mean",
    y = "Group",
    title = "Fitted vs Observed Group Means",
    subtitle = "Red X = observed, black = posterior median with 89% CI"
  ) +
  theme_rethinking()
```

### MCMC Diagnostics with bayesplot

```{r diagnostics}
# Trace plots
mcmc_trace(fit, pars = c("b_Intercept", "sigma", "sd_group__Intercept"))

# Pairs plot
mcmc_pairs(fit, pars = c("b_Intercept", "sigma", "sd_group__Intercept"))

# R-hat convergence diagnostic
rhat_values <- rhat(fit)
mcmc_rhat(rhat_values)

# Effective sample size
neff_values <- neff_ratio(fit)
mcmc_neff(neff_values)
```

### Posterior Predictive Checks

```{r pp-check}
# Compare observed to replicated data
pp_check(fit, type = "dens_overlay", ndraws = 50)
pp_check(fit, type = "stat", stat = "mean")
pp_check(fit, type = "stat_grouped", stat = "mean", group = "group")
```

### Shrinkage Analysis

Combine tidyrethinking utilities with brms output:

```{r shrinkage}
# Compute observed group statistics
group_stats <- ml_data %>%
  group_by(group) %>%
  summarise(
    mean_y = mean(y),
    n = n()
  )

# Extract between and within variance from fit
between_sd <- mean(posterior_draws$sd_group__Intercept)
within_sd <- mean(posterior_draws$sigma)

# Use tidyrethinking to compute shrinkage
shrinkage_data <- compute_shrinkage(
  group_means = group_stats$mean_y,
  group_ns = group_stats$n,
  within_sd = within_sd,
  between_sd = between_sd,
  grand_mean = mean(posterior_draws$b_Intercept)
)

# Visualize shrinkage
ggplot(shrinkage_data, aes(x = observed_mean, y = shrunken_mean)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(aes(size = n, color = shrinkage_factor), alpha = 0.7) +
  scale_color_viridis_c(option = "plasma") +
  scale_size_continuous(range = c(2, 6)) +
  labs(
    x = "Observed group mean",
    y = "Shrunken estimate (partial pooling)",
    color = "Shrinkage",
    size = "Group size",
    title = "Partial Pooling in Multilevel Model",
    subtitle = "Smaller groups shrink more toward grand mean"
  ) +
  theme_rethinking()
```

## Example: GLM with Prior Predictive Simulation

### Prior Predictive with tidyrethinking

```{r prior-pred}
# Use tidyrethinking for prior predictive
prior_samples <- generate_prior_predictive_linear(
  x = seq(-2, 2, length.out = 100),
  n_samples = 100,
  intercept_mean = 0,
  intercept_sd = 1,
  slope_mean = 0,
  slope_sd = 0.5,
  sigma = 1,
  seed = 42
)

# Visualize prior implications
ggplot(prior_samples, aes(x = x, y = y, group = sample)) +
  geom_line(alpha = 0.3, color = "steelblue") +
  labs(
    x = "Predictor",
    y = "Outcome",
    title = "Prior Predictive Distribution",
    subtitle = "What the model expects before seeing data"
  ) +
  theme_rethinking()
```

### Fit with brms

```{r glm-fit}
# Generate observed data
observed_data <- tibble(
  x = rnorm(100),
  y = 0.5 + 0.8 * x + rnorm(100, 0, 0.5)
)

# Fit with same priors
fit_glm <- brm(
  y ~ 1 + x,
  data = observed_data,
  prior = c(
    prior(normal(0, 1), class = Intercept),
    prior(normal(0, 0.5), class = b),
    prior(exponential(2), class = sigma)
  ),
  sample_prior = "yes",  # Save prior samples
  iter = 2000,
  chains = 4
)
```

### Compare Prior to Posterior with ggdist

```{r prior-post}
# Extract both prior and posterior
draws_df <- fit_glm %>%
  gather_draws(b_Intercept, b_x, sigma, regex = TRUE) %>%
  mutate(source = "Posterior")

prior_df <- prior_draws(fit_glm) %>%
  select(Intercept, b, sigma) %>%
  tidyr::pivot_longer(everything(), names_to = ".variable", values_to = ".value") %>%
  mutate(
    source = "Prior",
    .chain = 1,
    .iteration = row_number(),
    .draw = row_number()
  )

# Combine and plot
bind_rows(draws_df, prior_df) %>%
  ggplot(aes(x = .value, fill = source)) +
  stat_slab(alpha = 0.7) +
  facet_wrap(~.variable, scales = "free") +
  scale_fill_manual(values = c("Prior" = "grey70", "Posterior" = "steelblue")) +
  labs(
    x = "Parameter value",
    y = "Density",
    fill = "Distribution",
    title = "Prior vs Posterior Distributions"
  ) +
  theme_rethinking()
```

### Posterior Predictive Lines

```{r post-pred-lines}
# Use tidyrethinking to generate prediction data
new_x <- seq(min(observed_data$x), max(observed_data$x), length.out = 100)

# Extract posterior samples
post_samples <- as_draws_df(fit_glm) %>%
  select(b_Intercept, b_x) %>%
  slice_sample(n = 100)

# Generate lines
pred_lines <- generate_posterior_predictive_lines(
  x = new_x,
  intercepts = post_samples$b_Intercept,
  slopes = post_samples$b_x
)

# Plot with data
ggplot() +
  geom_line(
    data = pred_lines,
    aes(x = x, y = y, group = sample),
    alpha = 0.1,
    color = "steelblue"
  ) +
  geom_point(
    data = observed_data,
    aes(x = x, y = y),
    size = 2,
    alpha = 0.6
  ) +
  labs(
    x = "Predictor",
    y = "Outcome",
    title = "Posterior Predictive Distribution",
    subtitle = "100 plausible regression lines"
  ) +
  theme_rethinking()
```

## Integrating with posterior Package

```{r posterior-pkg}
# Convert to posterior's draws format
draws <- as_draws_rvars(fit)

# Summarize
summarise_draws(draws)

# Calculate derived quantities
draws_df <- as_draws_df(fit)
draws_df <- draws_df %>%
  mutate(
    # ICC using tidyrethinking formula
    tau_sq = sd_group__Intercept^2,
    sigma_sq = sigma^2,
    icc = tau_sq / (tau_sq + sigma_sq)
  )

# Visualize ICC
ggplot(draws_df, aes(x = icc)) +
  stat_halfeye(.width = c(0.66, 0.95)) +
  labs(
    x = "Intraclass Correlation (ICC)",
    y = "Density",
    title = "How much variance is between vs within groups?"
  ) +
  theme_rethinking()
```

## Best Practices

### 1. Start with Simulation

Always use tidyrethinking's simulation functions to:
- Understand causal structures
- Test your model on known data
- Validate recovery of parameters

```{r practice-sim, eval=FALSE}
# Simulate data with known parameters
sim_data <- simulate_confounding(
  n = 1000,
  effect_z_x = 1.0,
  effect_z_y = 1.5,
  effect_x_y = 0.5,
  seed = 42
)

# Fit model and check if you recover true effects
fit_test <- brm(y ~ x + z, data = sim_data)
# Compare estimates to true values: x->y = 0.5, z->y = 1.5
```

### 2. Use Prior Predictive Checks

```{r practice-prior, eval=FALSE}
# Visualize prior implications before fitting
prior_data <- generate_prior_predictive_linear(
  x = your_predictor_range,
  n_samples = 100,
  intercept_mean = 0,
  intercept_sd = 5,  # Adjust based on outcome scale
  slope_mean = 0,
  slope_sd = 1,
  sigma = 2
)
# Plot and ask: "Are these plausible relationships?"
```

### 3. Diagnose with Multiple Tools

Combine tidyrethinking, brms, and bayesplot:

```{r practice-diagnose, eval=FALSE}
# 1. Check R-hat with tidyrethinking
chains <- simulate_multiple_chains(n_chains = 4, n_steps = 1000)
compute_rhat(chains, var = "parameter")

# 2. Visual diagnostics with bayesplot
mcmc_trace(fit)
mcmc_dens_overlay(fit)

# 3. Posterior predictive checks
pp_check(fit, ndraws = 100)
```

### 4. Communicate Uncertainty

Use ggdist for publication-quality uncertainty visualization:

```{r practice-viz, eval=FALSE}
# Half-eye plots (density + interval)
posterior %>%
  ggplot(aes(x = parameter)) +
  stat_halfeye(.width = c(0.66, 0.95))

# Gradient interval plots
posterior %>%
  ggplot(aes(x = parameter)) +
  stat_gradientinterval()

# CCDF bar plots for discrete outcomes
posterior %>%
  ggplot(aes(x = count)) +
  stat_ccdfinterval()
```

## Resources

- **tidyrethinking**: Simulation and teaching tools
- **brms**: <https://paul-buerkner.github.io/brms/>
- **tidybayes**: <https://mjskay.github.io/tidybayes/>
- **ggdist**: <https://mjskay.github.io/ggdist/>
- **bayesplot**: <https://mc-stan.org/bayesplot/>
- **posterior**: <https://mc-stan.org/posterior/>

## Next Steps

- See `vignette("causal-inference")` for DAG-based workflows
- See `vignette("glm-models")` for binomial/Poisson models
- See `vignette("advanced-topics")` for measurement error, GPs, networks
