---
title: "Causal Inference with DAGs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Inference with DAGs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6
)
```

## Introduction

Causal inference requires careful thinking about which variables to condition on. This vignette demonstrates how to use **tidyrethinking** for:

- Creating and visualizing DAGs
- Simulating confounding, colliders, and mediation
- Understanding bias patterns
- Identifying valid adjustment sets

```{r setup}
library(tidyrethinking)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## DAG Fundamentals

### Creating DAG Layouts

tidyrethinking provides flexible DAG visualization tools:

```{r dag-layouts}
# Define nodes
nodes <- c("Z", "X", "Y")

# Horizontal layout (left to right)
horizontal <- create_dag_positions(nodes, layout = "horizontal")

# Vertical layout (top to bottom)
vertical <- create_dag_positions(nodes, layout = "vertical")

# Circular layout
circular <- create_dag_positions(nodes, layout = "circular")

# Custom positions
custom <- create_dag_positions(
  nodes,
  layout = "custom",
  positions = list(Z = c(0, 1), X = c(0, 0), Y = c(1, 0))
)

head(horizontal)
```

### Drawing DAGs

```{r dag-draw}
# Define a confounding structure: Z -> X, Z -> Y
dag_nodes <- create_dag_positions(
  nodes = c("Z", "X", "Y"),
  layout = "custom",
  positions = list(
    Z = c(0.5, 1),    # Confounder at top
    X = c(0, 0),      # Treatment at left
    Y = c(1, 0)       # Outcome at right
  )
)

# Create edges
dag_edges <- create_dag_edges(
  from = c("Z", "Z", "X"),
  to = c("X", "Y", "Y"),
  positions = dag_nodes
)

# Visualize
ggplot() +
  geom_segment(
    data = dag_edges,
    aes(x = x_start, y = y_start, xend = x_end, yend = y_end),
    arrow = arrow(length = unit(0.3, "cm"), type = "closed"),
    linewidth = 1,
    color = "steelblue"
  ) +
  geom_point(
    data = dag_nodes,
    aes(x = x, y = y),
    size = 20,
    color = "white",
    fill = "steelblue",
    shape = 21,
    stroke = 2
  ) +
  geom_text(
    data = dag_nodes,
    aes(x = x, y = y, label = node),
    size = 6,
    fontface = "bold"
  ) +
  coord_fixed(ratio = 1, xlim = c(-0.2, 1.2), ylim = c(-0.2, 1.2)) +
  theme_void() +
  labs(title = "Confounding Structure")
```

## Confounding

### Understanding the Bias

Confounding occurs when a common cause affects both treatment and outcome:

```{r confound-sim}
# Simulate confounded data
confounded <- simulate_confounding(
  n = 1000,
  effect_z_x = 1.0,   # Z influences X
  effect_z_y = 1.5,   # Z influences Y
  effect_x_y = 0.5,   # True causal effect
  seed = 42
)

head(confounded)
```

### Naive Analysis (Biased)

```{r confound-naive}
# Regression without controlling for confounder
naive_fit <- lm(y ~ x, data = confounded)
coef(naive_fit)["x"]  # Biased estimate

# The naive estimate conflates the true effect (0.5) 
# with the backdoor path through Z
```

### Correct Analysis

```{r confound-adjust}
# Regression controlling for confounder
adjusted_fit <- lm(y ~ x + z, data = confounded)
coef(adjusted_fit)["x"]  # Unbiased estimate ≈ 0.5

# Blocking the backdoor path recovers true effect
```

### Visualizing the Bias

```{r confound-viz}
# Create bins of confounder Z
confounded_binned <- confounded %>%
  mutate(z_bin = cut(z, breaks = 5))

# Plot relationship within Z bins
ggplot(confounded_binned, aes(x = x, y = y)) +
  geom_point(aes(color = z_bin), alpha = 0.4) +
  geom_smooth(aes(group = z_bin, color = z_bin), 
              method = "lm", se = FALSE, linewidth = 1) +
  geom_smooth(method = "lm", color = "black", linewidth = 1.5, 
              linetype = "dashed") +
  scale_color_viridis_d() +
  labs(
    x = "Treatment (X)",
    y = "Outcome (Y)",
    color = "Confounder\n(Z) bin",
    title = "Confounding Creates Spurious Association",
    subtitle = "Dashed line = naive estimate (biased)\nColored lines = within-Z slopes (unbiased)"
  ) +
  theme_rethinking()
```

## Collider Bias

### Understanding Selection Bias

A collider is a variable caused by both treatment and outcome. Conditioning on a collider creates spurious association:

```{r collider-sim}
# Simulate collider structure: X -> C <- Y
collider_data <- simulate_collider_bias(
  n = 1000,
  effect_x_c = 1.0,   # X causes C
  effect_y_c = 1.0,   # Y causes C
  effect_x_y = 0.0,   # No true effect!
  seed = 42
)

head(collider_data)
```

### Unconditional Analysis (Correct)

```{r collider-correct}
# X and Y are independent
cor(collider_data$x, collider_data$y)

# Regression without collider
unconditioned <- lm(y ~ x, data = collider_data)
coef(unconditioned)["x"]  # ≈ 0 (correct)
```

### Conditioned Analysis (Biased!)

```{r collider-biased}
# Condition on collider (selection bias)
conditioned <- lm(y ~ x + c, data = collider_data)
coef(conditioned)["x"]  # Non-zero (spurious!)

# Within-collider analysis
collider_subset <- collider_data %>%
  filter(c > quantile(c, 0.75))  # Select high C

cor(collider_subset$x, collider_subset$y)  # Spurious correlation!
```

### Visualizing Collider Bias

```{r collider-viz}
# Mark selection based on collider
collider_data <- collider_data %>%
  mutate(selected = c > quantile(c, 0.5))

ggplot(collider_data, aes(x = x, y = y)) +
  geom_point(aes(color = selected, alpha = selected), size = 2) +
  geom_smooth(
    data = filter(collider_data, selected),
    method = "lm",
    color = "red",
    linewidth = 1.5
  ) +
  geom_smooth(
    method = "lm",
    color = "black",
    linetype = "dashed",
    linewidth = 1.5
  ) +
  scale_color_manual(values = c("grey70", "coral")) +
  scale_alpha_manual(values = c(0.3, 0.8)) +
  labs(
    x = "X (independent of Y)",
    y = "Y (independent of X)",
    color = "Selected\n(high C)",
    alpha = "Selected\n(high C)",
    title = "Collider Bias Creates Spurious Negative Association",
    subtitle = "Red = selected sample (biased), Dashed = full sample (unbiased)"
  ) +
  theme_rethinking()
```

## Post-Treatment Bias (Mediation)

### Understanding Mediation

Post-treatment variables are caused by treatment and cause the outcome. Conditioning blocks the causal effect:

```{r mediation-sim}
# Simulate mediation: X -> M -> Y
mediation_data <- simulate_post_treatment_bias(
  n = 1000,
  effect_x_m = 0.7,   # X causes mediator M
  effect_m_y = 0.8,   # M causes Y
  direct_effect = 0.2,  # Direct X -> Y
  seed = 42
)

# Total effect = direct + indirect = 0.2 + (0.7 * 0.8) = 0.76
head(mediation_data)
```

### Total Effect (Correct for Total Causal Effect)

```{r mediation-total}
# Don't condition on mediator
total_effect <- lm(y ~ x, data = mediation_data)
coef(total_effect)["x"]  # ≈ 0.76 (total effect)
```

### Direct Effect Only

```{r mediation-direct}
# Condition on mediator to block indirect path
direct_effect <- lm(y ~ x + m, data = mediation_data)
coef(direct_effect)["x"]  # ≈ 0.2 (direct only)
coef(direct_effect)["m"]  # ≈ 0.8 (mediator effect)
```

### Mediation Analysis

```{r mediation-decompose}
# Calculate indirect effect
total_est <- coef(total_effect)["x"]
direct_est <- coef(direct_effect)["x"]
indirect_est <- total_est - direct_est

tibble(
  Effect = c("Total", "Direct", "Indirect", "Proportion mediated"),
  Estimate = c(
    total_est,
    direct_est,
    indirect_est,
    indirect_est / total_est
  )
)
```

### Visualizing Mediation

```{r mediation-viz}
# Create data for path visualization
mediation_paths <- tibble(
  path = c("X -> M -> Y", "X -> Y"),
  effect = c(indirect_est, direct_est),
  type = c("Indirect", "Direct")
)

ggplot(mediation_paths, aes(x = path, y = effect, fill = type)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c("Direct" = "steelblue", "Indirect" = "coral")) +
  labs(
    x = "Causal Path",
    y = "Effect Size",
    fill = "Effect Type",
    title = "Decomposing Total Causal Effect into Direct and Indirect",
    subtitle = sprintf("Total effect = %.2f (Direct: %.2f + Indirect: %.2f)",
                      total_est, direct_est, indirect_est)
  ) +
  theme_rethinking()
```

## Practical Guidelines

### 1. Draw Your DAG First

```{r guidelines-dag, eval=FALSE}
# Before analysis, explicitly state your causal assumptions
nodes <- c("Treatment", "Outcome", "Confounder", "Mediator", "Collider")

positions <- create_dag_positions(
  nodes,
  layout = "custom",
  positions = list(...)
)

edges <- create_dag_edges(
  from = c("Confounder", "Confounder", "Treatment", ...),
  to = c("Treatment", "Outcome", "Mediator", ...),
  positions = positions
)

# Visualize and discuss with collaborators
```

### 2. Simulate to Understand

```{r guidelines-sim, eval=FALSE}
# Generate data matching your DAG
sim_data <- simulate_confounding(n = 10000, ...)

# Try different analysis strategies
lm(outcome ~ treatment)                    # Biased?
lm(outcome ~ treatment + confounder)       # Unbiased?
lm(outcome ~ treatment + collider)         # Biased in other direction?

# Compare estimates to known truth
```

### 3. Test for Selection Bias

```{r guidelines-select, eval=FALSE}
# If you suspect collider bias:
full_data <- simulate_collider_bias(n = 1000)
selected_data <- full_data %>% filter(collider > threshold)

# Compare estimates
lm(y ~ x, data = full_data)      # Should be ≈ 0
lm(y ~ x, data = selected_data)  # Biased if collider
```

### 4. Identify Adjustment Sets

Rules of thumb:
- **Control for confounders**: Common causes of treatment and outcome
- **Don't control for colliders**: Common effects of treatment and outcome
- **Don't control for mediators**: Unless you want only direct effect
- **Control for pre-treatment variables**: Not post-treatment

## Complex DAGs

### Multiple Confounders

```{r complex-multiple}
# More realistic scenario
nodes <- c("Z1", "Z2", "X", "Y")
positions <- create_dag_positions(
  nodes,
  layout = "custom",
  positions = list(
    Z1 = c(0, 1),
    Z2 = c(1, 1),
    X = c(0.5, 0.5),
    Y = c(0.5, 0)
  )
)

# Z1 -> X, Z1 -> Y, Z2 -> X, Z2 -> Y, X -> Y
edges <- create_dag_edges(
  from = c("Z1", "Z1", "Z2", "Z2", "X"),
  to = c("X", "Y", "X", "Y", "Y"),
  positions = positions
)

# Need to adjust for both Z1 and Z2
```

### M-Bias

```{r complex-mbias}
# Spurious path through unobserved confounder
# U1 -> C <- U2, U1 -> X, U2 -> Y
# Controlling for C opens backdoor path!

nodes <- c("U1", "U2", "C", "X", "Y")
positions <- create_dag_positions(
  nodes,
  layout = "custom",
  positions = list(
    U1 = c(0, 1),
    U2 = c(1, 1),
    C = c(0.5, 0.8),
    X = c(0, 0),
    Y = c(1, 0)
  )
)

edges <- create_dag_edges(
  from = c("U1", "U2", "U1", "U2", "X"),
  to = c("C", "C", "X", "Y", "Y"),
  positions = positions
)

# Don't adjust for C even though it's correlated with both X and Y!
```

## Resources

- Pearl, J. (2009). *Causality: Models, Reasoning, and Inference*
- Hernán & Robins (2020). *Causal Inference: What If*
- McElreath, R. (2020). *Statistical Rethinking*, Chapters 5-6
- **dagitty**: <http://dagitty.net/> for identifying adjustment sets
- **ggdag**: <https://ggdag.malco.io/> for more DAG visualization

## Next Steps

- See `vignette("bayesian-workflow")` for fitting causal models with brms
- See `vignette("glm-models")` for causal inference with GLMs
- See `vignette("advanced-topics")` for instrumental variables
