---
title: "Getting Started with tidyrethinking"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tidyrethinking}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

The **tidyrethinking** package provides modern, tidyverse-aligned tools for learning and applying Bayesian statistics through the lens of Richard McElreath's Statistical Rethinking course. This vignette introduces the core concepts and demonstrates basic usage.

```{r setup}
library(tidyrethinking)
library(dplyr)
library(ggplot2)
```

## Garden of Forking Paths

The garden of forking paths is a metaphor for Bayesian updating. Each path represents a possible hypothesis, and observations eliminate paths that are inconsistent with the data.

### Simulating the Garden

Let's simulate three observations: water (1), land (0), water (1). We have four marbles: one land (0) and three water (1).

```{r garden-simulate}
observations <- c(1, 0, 1)
possibilities <- c(0, 1, 1, 1)

paths <- simulate_garden_paths(observations, possibilities)
head(paths, 10)
```

### Visualizing Paths

We can create a radial layout to visualize how paths branch and which remain viable:

```{r garden-layout}
paths_layout <- create_garden_layout(paths)

ggplot(paths_layout, aes(x = x, y = y, group = path_id)) +
  geom_path(aes(alpha = viable, color = viable)) +
  geom_point(aes(color = viable, size = viable)) +
  scale_alpha_manual(values = c(0.2, 1)) +
  scale_color_manual(values = c("grey50", "steelblue")) +
  scale_size_manual(values = c(1, 2)) +
  coord_fixed() +
  theme_rethinking() +
  theme(legend.position = "bottom") +
  labs(title = "Garden of Forking Paths",
       subtitle = "Observations: Water, Land, Water")
```

### Counting Outcomes

We can count the number of ways each outcome can occur:

```{r garden-count}
count_path_outcomes(observations, possibilities)
```

## Globe Tossing

The globe tossing example demonstrates Bayesian updating with a conjugate Beta-Binomial model.

### Simulating Tosses

First, let's generate some simulated globe tosses:

```{r globe-simulate}
tosses <- simulate_globe_tosses(
  n = 9, 
  prob_water = 0.7, 
  seed = 42
)

tosses
```

### Bayesian Updating

Now let's compute the posterior distribution after each observation:

```{r globe-update}
posteriors <- compute_beta_updates(
  observations = tosses$result,
  prior_alpha = 1,
  prior_beta = 1
)

posteriors
```

### Visualizing Posterior Evolution

We can see how the posterior distribution evolves with each new observation:

```{r globe-density}
density_data <- generate_posterior_density(posteriors, n_points = 100)

ggplot(density_data, aes(x = p, y = density, group = toss)) +
  geom_line(aes(color = toss), linewidth = 1) +
  scale_color_viridis_c() +
  labs(
    x = "Proportion water",
    y = "Posterior density",
    color = "Toss",
    title = "Posterior Evolution with Each Observation"
  ) +
  theme_rethinking()
```

### Posterior Intervals

Calculate credible intervals from the final posterior:

```{r globe-interval}
final_posterior <- posteriors[nrow(posteriors), ]

interval <- compute_posterior_interval(
  alpha = final_posterior$alpha,
  beta = final_posterior$beta,
  prob = 0.89
)

interval
```

### Posterior Predictive Simulation

We can simulate future observations using the posterior predictive distribution:

```{r globe-predictive}
predictive_samples <- sample_posterior_predictive(
  posterior_alpha = final_posterior$alpha,
  posterior_beta = final_posterior$beta,
  n_tosses = 9,
  n_samples = 1000,
  seed = 42
)

ggplot(predictive_samples, aes(x = n_water)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  geom_vline(xintercept = sum(tosses$result == "W"), 
             linetype = "dashed", color = "red", linewidth = 1) +
  labs(
    x = "Number of water observations (out of 9)",
    y = "Count",
    title = "Posterior Predictive Distribution",
    subtitle = "Red line shows observed data"
  ) +
  theme_rethinking()
```

## Central Limit Theorem via Random Walk

The random walk simulation demonstrates how sums of random variables converge to normal distributions.

### Simulating Random Walk

```{r walk-simulate}
walk_data <- simulate_random_walk(
  n_individuals = 1000,
  n_steps = 16,
  seed = 42
)

# View a few paths
walk_data %>%
  filter(individual <= 5) %>%
  ggplot(aes(x = step, y = position, color = factor(individual))) +
  geom_line() +
  geom_point() +
  labs(
    x = "Step",
    y = "Position",
    color = "Individual",
    title = "Five Random Walk Paths"
  ) +
  theme_rethinking()
```

### Final Distribution

After 16 steps, the distribution of positions approximates a normal distribution:

```{r walk-final}
final_positions <- walk_data %>%
  filter(step == max(step))

ggplot(final_positions, aes(x = position)) +
  geom_histogram(aes(y = ..density..), binwidth = 1, 
                 fill = "steelblue", alpha = 0.7) +
  stat_function(
    fun = dnorm,
    args = list(
      mean = mean(final_positions$position),
      sd = sd(final_positions$position)
    ),
    color = "red",
    linewidth = 1
  ) +
  labs(
    x = "Position",
    y = "Density",
    title = "Final Distribution after 16 Steps",
    subtitle = "Red curve shows matching normal distribution"
  ) +
  theme_rethinking()
```

### Comparing to Normal

We can formally compare the empirical distribution to the theoretical normal:

```{r walk-compare}
comparison <- compare_to_normal(walk_data, step_to_compare = 16)

ggplot(comparison, aes(x = x)) +
  geom_line(aes(y = empirical_density, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = theoretical_density, color = "Normal"), 
            linetype = "dashed", linewidth = 1) +
  scale_color_manual(values = c("Observed" = "steelblue", "Normal" = "red")) +
  labs(
    x = "Position",
    y = "Density",
    color = "Distribution",
    title = "Empirical vs. Theoretical Normal Distribution"
  ) +
  theme_rethinking()
```

### Testing Normality

```{r walk-test}
normality_test <- test_walk_normality(walk_data, final_step_only = TRUE)
normality_test
```

The high Shapiro-Wilk p-value and low skewness/kurtosis indicate the distribution is approximately normal.

## Next Steps

This vignette introduced the core functionality of tidyrethinking for Chapters 2-3. Additional vignettes cover:

- Linear models and regression (Chapter 4)
- Causal inference with DAGs (Chapters 5-6)
- MCMC and model comparison (Chapters 7-8)
- Generalized linear models (Chapters 9-11)
- Multilevel models (Chapters 12-14)
- Advanced topics (Chapters 15-19)

For more information, see the package documentation:

```{r, eval=FALSE}
?tidyrethinking
help(package = "tidyrethinking")
```
