---
title: "Generalized Linear Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized Linear Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette covers generalized linear models (GLMs) for non-normal outcomes:

- **Binomial models**: Binary and count/trial data
- **Poisson models**: Count data
- **Zero-inflated models**: Excess zeros

```{r setup}
library(tidyrethinking)
library(dplyr)
library(ggplot2)
library(tidyr)
```

## Binomial Models

### Logit Link Function

The logit link maps probabilities to the real line:

```{r logit}
# Probability to log-odds
probs <- seq(0.01, 0.99, by = 0.01)
log_odds <- logit(probs)

tibble(prob = probs, log_odds = log_odds) %>%
  ggplot(aes(x = prob, y = log_odds)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(
    x = "Probability",
    y = "Log-odds",
    title = "Logit Link Function"
  ) +
  theme_rethinking()

# Log-odds to probability
log_odds_seq <- seq(-4, 4, by = 0.1)
probs_back <- inv_logit(log_odds_seq)

tibble(log_odds = log_odds_seq, prob = probs_back) %>%
  ggplot(aes(x = log_odds, y = prob)) +
  geom_line(color = "coral", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey50") +
  labs(
    x = "Log-odds",
    y = "Probability",
    title = "Inverse Logit (Logistic) Function"
  ) +
  theme_rethinking()
```

### Simulating Binomial Data

```{r binomial-sim}
# Single predictor
set.seed(42)
binom_data <- simulate_binomial_data(
  n = 200,
  intercept = -1,    # log-odds at x = 0
  slopes = 0.8,      # effect on log-odds scale
  seed = 42
)

head(binom_data)

# Visualize
ggplot(binom_data, aes(x = x_1, y = y)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "glm", method.args = list(family = "binomial"),
              color = "steelblue", linewidth = 1) +
  labs(
    x = "Predictor",
    y = "Binary outcome",
    title = "Simulated Binomial Data"
  ) +
  theme_rethinking()
```

### Grid Approximation for Posterior

```{r binomial-grid}
# Simple example: 6 successes in 9 trials
posterior_samples <- sample_binomial_posterior(
  y = 6,
  n = 9,
  x_matrix = matrix(1, nrow = 1),  # Intercept only
  n_samples = 1000,
  seed = 42
)

# Convert to probability scale
posterior_samples <- posterior_samples %>%
  mutate(prob = inv_logit(intercept))

# Visualize posterior
ggplot(posterior_samples, aes(x = prob)) +
  geom_histogram(aes(y = ..density..), bins = 30, 
                 fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 6/9, color = "red", 
             linetype = "dashed", linewidth = 1) +
  labs(
    x = "Probability of success",
    y = "Posterior density",
    title = "Posterior Distribution (6/9 successes)",
    subtitle = "Red line = observed proportion"
  ) +
  theme_rethinking()
```

### Credible Intervals

```{r binomial-interval}
intervals <- compute_binomial_intervals(
  posterior_samples$intercept,
  probs = c(0.50, 0.89, 0.95)
)

intervals
```

### Effect Sizes: Log-Odds Ratio

```{r binomial-or}
# Simulate data with two predictors
multi_data <- simulate_binomial_data(
  n = 500,
  intercept = -0.5,
  slopes = c(0.7, -0.4),
  seed = 42
)

# Fit model
fit <- glm(y ~ x_1 + x_2, data = multi_data, family = binomial())

# Log-odds ratios
coef(fit)

# Convert to odds ratios (exponentiate)
exp(coef(fit))

# Interpretation: 
# - exp(β) > 1: predictor increases odds
# - exp(β) < 1: predictor decreases odds
# - exp(β) = 1: no effect

# Compare to simulation truth
compute_log_odds_ratio(
  p_baseline = inv_logit(-0.5),
  p_treatment = inv_logit(-0.5 + 0.7)
)
```

### Visualizing Predictions

```{r binomial-predict}
# Create prediction grid
pred_grid <- expand_grid(
  x_1 = seq(min(multi_data$x_1), max(multi_data$x_1), length.out = 50),
  x_2 = quantile(multi_data$x_2, c(0.25, 0.5, 0.75))
)

# Add predictions
pred_grid$prob <- predict(fit, newdata = pred_grid, type = "response")

# Plot
ggplot(pred_grid, aes(x = x_1, y = prob, color = factor(x_2))) +
  geom_line(linewidth = 1) +
  geom_point(data = multi_data, aes(y = y, color = NULL), 
             alpha = 0.2, inherit.aes = TRUE) +
  scale_color_viridis_d(option = "plasma") +
  labs(
    x = "Predictor 1",
    y = "Probability of success",
    color = "Predictor 2",
    title = "Predicted Probabilities from Binomial Model"
  ) +
  theme_rethinking()
```

## Poisson Models

### Log Link Function

The log link ensures positive predictions for counts:

```{r poisson-link}
# Linear predictor to expected count
linear_pred <- seq(-2, 3, by = 0.1)
expected_count <- exp(linear_pred)

tibble(linear_pred, expected_count) %>%
  ggplot(aes(x = linear_pred, y = expected_count)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  labs(
    x = "Linear predictor (log scale)",
    y = "Expected count",
    title = "Exponential (Inverse Log) Link Function"
  ) +
  theme_rethinking()
```

### Simulating Poisson Data

```{r poisson-sim}
poisson_data <- simulate_poisson_data(
  n = 300,
  intercept = 1.5,      # log(count) at x = 0
  slopes = c(0.3, -0.2),
  seed = 42
)

head(poisson_data)

# Visualize
ggplot(poisson_data, aes(x = x_1, y = y)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_smooth(method = "glm", method.args = list(family = "poisson"),
              color = "darkgreen", linewidth = 1) +
  labs(
    x = "Predictor",
    y = "Count",
    title = "Simulated Poisson Data"
  ) +
  theme_rethinking()
```

### Rate Ratios

```{r poisson-rate}
# Fit model
pois_fit <- glm(y ~ x_1 + x_2, data = poisson_data, family = poisson())

# Coefficients are on log scale
coef(pois_fit)

# Convert to rate ratios
rate_ratios <- compute_rate_ratio(coef(pois_fit)[-1])  # Exclude intercept
rate_ratios

# Interpretation:
# - Rate ratio > 1: predictor increases count
# - Rate ratio < 1: predictor decreases count
# - Rate ratio = 1: no effect
```

### Overdispersion

Poisson models assume mean = variance. Real data often have extra variance:

```{r poisson-dispersion}
# Check for overdispersion
tibble(
  Parameter = c("Residual deviance", "df", "Dispersion"),
  Value = c(
    deviance(pois_fit),
    df.residual(pois_fit),
    deviance(pois_fit) / df.residual(pois_fit)
  )
)

# Dispersion ≈ 1: good fit
# Dispersion > 1: overdispersion (consider negative binomial)
# Dispersion < 1: underdispersion (rare)
```

## Zero-Inflated Poisson

### Understanding Excess Zeros

Many count datasets have more zeros than Poisson predicts:

```{r zip-concept}
# Compare Poisson to ZIP
n_sim <- 1000
lambda <- 2
zero_prob <- 0.3

# Standard Poisson
pois_counts <- rpois(n_sim, lambda)

# Zero-inflated Poisson
zip_data <- simulate_zero_inflated_poisson(
  n = n_sim,
  intercept = log(lambda),
  slopes = NULL,
  zero_prob = zero_prob,
  seed = 42
)

# Compare distributions
compare_df <- tibble(
  count = 0:10,
  Poisson = dpois(0:10, lambda),
  ZIP = sapply(0:10, function(k) {
    ifelse(k == 0, 
           zero_prob + (1 - zero_prob) * dpois(0, lambda),
           (1 - zero_prob) * dpois(k, lambda))
  })
) %>%
  pivot_longer(-count, names_to = "Model", values_to = "Probability")

ggplot(compare_df, aes(x = count, y = Probability, fill = Model)) +
  geom_col(position = "dodge", alpha = 0.7) +
  scale_fill_manual(values = c("Poisson" = "steelblue", "ZIP" = "coral")) +
  labs(
    x = "Count",
    y = "Probability",
    title = "Zero-Inflated Poisson vs Standard Poisson",
    subtitle = sprintf("λ = %.1f, P(structural zero) = %.1f", lambda, zero_prob)
  ) +
  theme_rethinking()
```

### Simulating ZIP Data

```{r zip-sim}
# Simulate data with predictors
zip_full <- simulate_zero_inflated_poisson(
  n = 500,
  intercept = 1.0,
  slopes = c(0.5, -0.3),
  zero_prob = 0.4,
  seed = 42
)

# Check proportion of zeros
mean(zip_full$y == 0)  # ~40% zeros
mean(rpois(500, exp(1.0)) == 0)  # Poisson would predict ~37% zeros
```

### Two-Part Models

ZIP models have two parts:
1. **Zero-inflation**: Logistic model for structural zeros
2. **Count model**: Poisson for non-structural counts

```{r zip-interpret}
# In practice, use packages like pscl or brms for ZIP
# Here we show the intuition

# Part 1: Which observations are structural zeros?
zip_full <- zip_full %>%
  mutate(
    structural_zero = y == 0 & runif(n()) < 0.4,
    count_process = ifelse(structural_zero, NA, y)
  )

# Visualize two processes
ggplot(zip_full, aes(x = x_1, y = y, color = structural_zero)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_color_manual(
    values = c("FALSE" = "darkgreen", "TRUE" = "grey70"),
    labels = c("Count process", "Structural zero")
  ) +
  labs(
    x = "Predictor",
    y = "Count",
    color = "Type",
    title = "Zero-Inflated Poisson Data",
    subtitle = "Two different sources of zeros"
  ) +
  theme_rethinking()
```

## Model Comparison

### When to Use Each Model

```{r model-comparison}
tibble(
  Outcome = c("Binary", "Count (0-n)", "Count with trials", 
              "Count (rare)", "Count (excess zeros)"),
  Model = c("Binomial (logit)", "Poisson", "Binomial", 
            "Poisson", "Zero-inflated Poisson"),
  Link = c("logit", "log", "logit", "log", "logit + log"),
  Example = c(
    "Survived (yes/no)",
    "Number of events",
    "Successes out of trials",
    "Disease cases per year",
    "Fish caught (many zeros)"
  )
) %>%
  knitr::kable()
```

### Simulating from Different Models

```{r model-compare-sim}
set.seed(42)
n <- 200
x <- rnorm(n)

# Different data generating processes
comparison_data <- tibble(
  x = x,
  Binary = rbinom(n, 1, prob = inv_logit(-0.5 + 0.8 * x)),
  Binomial = rbinom(n, 10, prob = inv_logit(-0.5 + 0.8 * x)),
  Poisson = rpois(n, lambda = exp(1 + 0.5 * x)),
  ZIP = simulate_zero_inflated_poisson(
    n, intercept = 1, slopes = 0.5, zero_prob = 0.3
  )$y
)

# Visualize distributions
comparison_long <- comparison_data %>%
  pivot_longer(-x, names_to = "Model", values_to = "y")

ggplot(comparison_long, aes(x = y)) +
  geom_histogram(binwidth = 1, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~Model, scales = "free") +
  labs(
    x = "Outcome value",
    y = "Count",
    title = "Different GLM Outcome Distributions"
  ) +
  theme_rethinking()
```

## Practical Tips

### 1. Check Your Scale

```{r tips-scale}
# Probabilities: use logit link
# Counts: use log link
# Always check that predictions make sense

# Bad: predicting negative probabilities
# Good: predictions bounded between 0 and 1

# Bad: predicting negative counts
# Good: predictions always positive
```

### 2. Interpret on Natural Scale

```{r tips-interpret}
# Coefficients are on link scale
# Transform for interpretation

# Binomial: exp(β) = odds ratio
beta_logit <- 0.5
sprintf("1 unit increase multiplies odds by %.2f", exp(beta_logit))

# Poisson: exp(β) = rate ratio
beta_log <- 0.3
sprintf("1 unit increase multiplies count by %.2f", exp(beta_log))
```

### 3. Use Prior Predictive Checks

```{r tips-prior, eval=FALSE}
# Simulate from your priors before fitting
prior_samples <- generate_prior_predictive_linear(
  x = your_predictors,
  intercept_mean = 0,
  intercept_sd = 1.5,  # On log-odds scale
  slope_mean = 0,
  slope_sd = 1,
  sigma = 0  # No residual for binomial
)

# Do predictions look reasonable?
# Are they in the right range?
```

### 4. Check Predictions

```{r tips-check, eval=FALSE}
# Posterior predictive checks
fitted_values <- predict(model, type = "response")

# Compare observed to expected
tibble(observed = y, fitted = fitted_values) %>%
  ggplot(aes(x = fitted, y = observed)) +
  geom_point(alpha = 0.3) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed")
```

## Resources

- McElreath, R. (2020). *Statistical Rethinking*, Chapters 9-11
- Gelman & Hill (2007). *Data Analysis Using Regression and Multilevel/Hierarchical Models*
- **brms** for Bayesian GLMs: <https://paul-buerkner.github.io/brms/>
- **pscl** for zero-inflated models: <https://github.com/atahk/pscl>

## Next Steps

- See `vignette("bayesian-workflow")` for fitting GLMs with brms
- See `vignette("multilevel-models")` for nested and crossed structures
- See `vignette("advanced-topics")` for robust regression and measurement error
