---
title: "MCMC demonstrations with tidyrethinking"
description: "Metropolis sampler, target/gradient functions, and diagnostics using tidyverse-friendly APIs"
vignette: >
  %\VignetteIndexEntry{MCMC demonstrations with tidyrethinking}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
    number-sections: false
---

This vignette demonstrates MCMC building blocks provided by tidyrethinking: target densities, analytic gradients, a Metropolis sampler, and simple diagnostics, all with tidyverse-friendly, data-first APIs.

## Setup

```{r}
library(tidyrethinking)
library(dplyr)
library(ggplot2)
```

## 1. Targets and gradients (Normal with unknown mean/sigma)

```{r}
dat <- tibble::tibble(y = rnorm(200, 0, 1))

# Negative log posterior (energy) and gradient at a point
U  <- normal_mu_logsigma_target(dat, c(0, 0))
G  <- normal_mu_logsigma_gradient(dat, c(0, 0))
U; G
```

Vectorized across parameter rows:

```{r}
params <- tibble::tibble(mu = c(-0.5, 0, 0.5), log_sigma = c(-0.2, 0, 0.2))
normal_mu_logsigma_target(dat, params)
```

## 2. Metropolis sampler

```{r}
set.seed(1)
draws <- metropolis_sampler(
  target_fn   = normal_mu_logsigma_target,
  data        = dat,
  init        = c(0, 0),
  n_samples   = 2000,
  step        = c(0.05, 0.05),
  chains      = 2
)

# Discard burn-in
post <- dplyr::filter(draws, iter > 500)
```

Inspect the marginal for `mu`:

```{r}
post |> ggplot(aes(x = mu)) +
  geom_histogram(bins = 40, color = "white", fill = "steelblue") +
  theme_minimal() +
  labs(title = "Posterior draws for mu")
```

## 3. Diagnostics (R-hat and ESS)

```{r}
rhat_mu <- mcmc_rhat(post, "mu")
ess_mu  <- mcmc_ess(post,  "mu")
list(rhat_mu = rhat_mu, ess_mu = ess_mu)
```

## 4. Parallel chains with purrr::in_parallel (optional)

If you enable parallel processing via `mirai::daemons(n)`, mapping over chains can be parallelized internally. Parallelization is optional and controlled by the user; without daemons set, everything runs sequentially.

```{r eval=FALSE}
if (interactive() && rlang::is_installed("mirai") && rlang::is_installed("carrier")) {
  mirai::daemons(4)
  draws_par <- metropolis_sampler(
    target_fn   = normal_mu_logsigma_target,
    data        = dat,
    init        = c(0, 0),
    n_samples   = 5000,
    step        = c(0.05, 0.05),
    chains      = 4
  )
  mirai::daemons(0)
}
```

