---
title: "tidyrethinking Package Overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidyrethinking Package Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## What is tidyrethinking?

**tidyrethinking** modernizes Statistical Rethinking course materials into a production-ready R package. It provides:

- ✅ **85+ functions** organized by topic (Chapters 2-18)
- ✅ **Tidyverse-aligned** API (tibbles, pipes, snake_case)
- ✅ **Comprehensive documentation** with examples
- ✅ **Full test coverage** ensuring reliability
- ✅ **Reproducible workflows** via seed parameters
- ✅ **Integration-ready** with brms, tidybayes, ggdist

```{r setup}
library(tidyrethinking)
library(dplyr)
library(ggplot2)
```

## Package Structure

### Foundation Utilities

Core tools used across all chapters:

```{r foundation}
# Geometry transformations
polar_to_cartesian(r = 1, theta = pi/4)
cartesian_to_polar(x = 1, y = 1)

# Plotting themes
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  theme_rethinking()

# Color palettes
rethinking_colors(3)

# Validation helpers (used internally)
# validate_probability(0.5)  # Ensures 0 ≤ p ≤ 1
# validate_positive(10)       # Ensures x > 0
```

### Chapter 2: Bayesian Fundamentals

```{r ch2, eval=FALSE}
# Garden of forking paths
paths <- simulate_garden_paths(c(1, 0, 1), c(0, 1, 1, 1))
layout <- create_garden_layout(paths)
count_path_outcomes(c(1, 0, 1), c(0, 1, 1, 1))

# Globe tossing
tosses <- simulate_globe_tosses(n = 9, prob_water = 0.7)
posteriors <- compute_beta_updates(tosses$result)
intervals <- compute_posterior_interval(alpha = 10, beta = 4, prob = 0.89)
predictive <- sample_posterior_predictive(10, 4, n_tosses = 9, n_samples = 1000)
```

### Chapters 3-4: Gaussian Models

```{r ch3-4, eval=FALSE}
# Random walk (CLT demonstration)
walk <- simulate_random_walk(n_individuals = 1000, n_steps = 16)
comparison <- compare_to_normal(walk, step_to_compare = 16)
test <- test_walk_normality(walk)

# Prior/posterior predictive
prior_pred <- generate_prior_predictive_linear(x, n_samples = 100)
posterior <- compute_linear_posterior(x, y, prior_mean, prior_sd, sigma)
post_samples <- sample_linear_posterior(posterior, n_samples = 1000)
pred_lines <- generate_posterior_predictive_lines(x, intercepts, slopes)
intervals <- compute_prediction_intervals(predictions, probs = c(0.67, 0.89, 0.97))
```

### Chapters 5-6: Causal Inference

```{r ch5-6, eval=FALSE}
# DAG utilities
positions <- create_dag_positions(c("Z", "X", "Y"), layout = "horizontal")
edges <- create_dag_edges(from = c("Z", "Z"), to = c("X", "Y"), positions)

# Causal simulations
confounded <- simulate_confounding(n = 1000, effect_z_x = 1, effect_z_y = 1.5)
collider <- simulate_collider_bias(n = 1000, effect_x_c = 1, effect_y_c = 1)
mediation <- simulate_post_treatment_bias(n = 1000, effect_x_m = 0.7, effect_m_y = 0.8)
```

### Chapter 8: MCMC

```{r ch8, eval=FALSE}
# Metropolis algorithm
mcmc <- simulate_metropolis(n_steps = 10000, start = c(0, 0))
diagnostics <- compute_mcmc_diagnostics(mcmc)

# Multiple chains
chains <- simulate_multiple_chains(n_chains = 4, n_steps = 1000)
rhat <- compute_rhat(chains, var = "island")

# Visualization prep
trace_data <- create_trace_plot_data(chains)
```

### Chapters 9-10: Binomial GLMs

```{r ch9-10, eval=FALSE}
# Simulate binomial data
data <- simulate_binomial_data(n = 200, intercept = -1, slopes = c(0.8, -0.5))

# Link functions
log_odds <- logit(0.7)
prob <- inv_logit(0.5)

# Posterior inference
posterior <- sample_binomial_posterior(y, n, x_matrix, n_samples = 1000)
intervals <- compute_binomial_intervals(posterior$intercept, probs = c(0.89, 0.95))

# Effect sizes
odds_ratio <- compute_log_odds_ratio(p_baseline = 0.3, p_treatment = 0.6)
```

### Chapter 11: Poisson GLMs

```{r ch11, eval=FALSE}
# Simulate count data
counts <- simulate_poisson_data(n = 300, intercept = 1.5, slopes = 0.3)

# Zero-inflated
zip_data <- simulate_zero_inflated_poisson(n = 500, intercept = 1, slopes = 0.5, zero_prob = 0.4)

# Rate ratios
rate_ratio <- compute_rate_ratio(coefficients)

# Posterior
posterior <- sample_poisson_posterior(y, x_matrix, n_samples = 1000)
intervals <- compute_poisson_intervals(posterior$intercept)
```

### Chapters 12-13: Multilevel Basics

```{r ch12-13, eval=FALSE}
# Varying intercepts
vi_data <- simulate_varying_intercepts(n_groups = 20, n_per_group = 10)

# Varying slopes
vs_data <- simulate_varying_slopes(n_groups = 15, n_per_group = 20)

# Correlated effects
corr_data <- simulate_correlated_effects(n_groups = 20, n_per_group = 15, correlation = 0.6)

# Shrinkage
shrinkage <- compute_shrinkage(group_means, group_ns, within_sd = 2, between_sd = 1)

# ICC
icc <- compute_icc(between_group_sd = 2, within_group_sd = 1)
```

### Chapter 14: Advanced Multilevel

```{r ch14, eval=FALSE}
# Cross-classified
cross <- simulate_cross_classified(n_groups_1 = 10, n_groups_2 = 8, n_per_combo = 5)

# Temporal autocorrelation
temporal <- simulate_temporal_autocorrelation(n_groups = 5, n_timepoints = 50, rho = 0.7)

# Spatial autocorrelation
spatial <- simulate_spatial_autocorrelation(n_locations = 50, rho = 0.8)

# Post-stratification
weights <- compute_poststrat_weights(sample_counts, population_counts)
```

### Chapter 15: Measurement Error

```{r ch15, eval=FALSE}
# Measurement error
error_data <- simulate_measurement_error(n = 500, true_effect = 0.8, measurement_sd = 0.5)
bias <- compute_attenuation_bias(true_effect = 0.8, true_sd = 1, measurement_sd = 0.5)

# Missing data patterns
mcar <- simulate_mcar(n = 500, true_effect = 0.5, missing_prob = 0.3)
mar <- simulate_mar(n = 500, true_effect = 0.5, covariate_effect_on_missing = 1.5)

# Imputation diagnostics
mice_summary <- compute_mice_summary(imputed_datasets)
```

### Chapter 16: Gaussian Processes

```{r ch16, eval=FALSE}
# Kernels
cov_se <- compute_squared_exponential(x1, x2, eta_sq = 1, rho_sq = 1)
cov_periodic <- compute_periodic_covariance(x1, x2, eta_sq = 1, rho_sq = 0.5, period = 1)

# Prior samples
prior <- sample_gp_prior(x = seq(-3, 3, 0.1), n_samples = 20)

# Posterior
posterior <- compute_gp_posterior(x_obs, y_obs, x_pred, eta_sq = 1, rho_sq = 1)
post_samples <- sample_gp_posterior(x_obs, y_obs, x_pred, n_samples = 100)
```

### Chapter 17: Social Networks

```{r ch17, eval=FALSE}
# Dyadic network
network <- simulate_dyadic_network(n_nodes = 50, reciprocity = 2)
adj_matrix <- create_adjacency_matrix(network, value_col = "tie")
stats <- compute_network_stats(network)

# Transitive closure
trans_net <- simulate_transitive_closure(n_nodes = 30, initial_ties = 40, transitivity_prob = 0.7)

# Independence baseline
indep <- compute_dyadic_independence(network)
```

### Chapter 18: Advanced Missing Data

```{r ch18, eval=FALSE}
# MNAR
mnar <- simulate_mnar(n = 500, true_effect = 0.5, outcome_effect_on_missing = 1.0)
bias <- compute_selection_bias(complete_estimate, full_estimate)

# Instrumental variables
iv_data <- simulate_instrument(n = 1000, instrument_strength = 0.7, true_effect = 0.5)
tsls <- compute_tsls(iv_data, y_var = "y", x_var = "x", z_var = "z")

# Censored data
censored <- simulate_censored(n = 500, true_mean = 5, lower_limit = 0, upper_limit = 10)
summary <- compute_censoring_summary(censored)
```

## Integration with Bayesian Ecosystem

### With brms

```{r brms, eval=FALSE}
library(brms)
library(tidybayes)
library(ggdist)

# Simulate with tidyrethinking
data <- simulate_varying_intercepts(n_groups = 20, n_per_group = 10)

# Fit with brms
fit <- brm(y ~ 1 + (1 | group), data = data)

# Extract with tidybayes
posterior <- fit %>% spread_draws(b_Intercept, sd_group__Intercept)

# Visualize with ggdist
posterior %>%
  ggplot(aes(x = sd_group__Intercept)) +
  stat_halfeye() +
  theme_rethinking()

# Analyze with tidyrethinking
shrinkage <- compute_shrinkage(
  group_means = ...,
  group_ns = ...,
  within_sd = mean(posterior$sigma),
  between_sd = mean(posterior$sd_group__Intercept)
)
```

### With cmdstanr

```{r cmdstan, eval=FALSE}
library(cmdstanr)

# Simulate data
data <- simulate_binomial_data(n = 100, intercept = -1, slopes = 0.5)

# Prepare for Stan
stan_data <- list(
  N = nrow(data),
  y = data$y,
  x = data$x_1
)

# Fit model
model <- cmdstan_model("logistic_regression.stan")
fit <- model$sample(data = stan_data)

# Diagnose with tidyrethinking
chains <- fit$draws(format = "df")
rhat <- compute_rhat(chains, var = "beta")
```

### With ggdist and bayesplot

```{r viz, eval=FALSE}
library(ggdist)
library(bayesplot)

# Generate predictions from tidyrethinking
posterior_lines <- generate_posterior_predictive_lines(
  x = seq(-2, 2, length.out = 100),
  intercepts = posterior$intercept,
  slopes = posterior$slope
)

# Visualize with ggdist
ggplot(posterior_lines, aes(x = x, y = y)) +
  stat_lineribbon(aes(group = sample), .width = c(0.50, 0.89, 0.95)) +
  scale_fill_brewer() +
  theme_rethinking()

# MCMC diagnostics with bayesplot
mcmc_trace(fit)
mcmc_dens_overlay(fit)
```

## Design Philosophy

### 1. Tidyverse Principles

```{r tidy-principles, eval=FALSE}
# ✅ Returns tibbles
result <- simulate_garden_paths(c(1, 0, 1), c(0, 1, 1, 1))
class(result)  # "tbl_df" "tbl" "data.frame"

# ✅ Pipe-friendly
data %>%
  simulate_confounding(n = 1000) %>%
  mutate(outcome_group = cut(y, breaks = 3)) %>%
  ggplot(aes(x = x, y = y, color = outcome_group)) +
  geom_point()

# ✅ Snake_case naming
simulate_varying_intercepts()  # Not SimulateVaryingIntercepts()
compute_shrinkage()            # Not ComputeShrinkage()

# ✅ Explicit arguments (no abbreviations)
simulate_garden_paths(
  observations = c(1, 0, 1),
  possibilities = c(0, 1, 1, 1)
)
```

### 2. Reproducibility

```{r reproducibility, eval=FALSE}
# All stochastic functions accept seed
set.seed(42)
data1 <- simulate_binomial_data(n = 100, intercept = 0, slopes = 0.5)

# Equivalent
data2 <- simulate_binomial_data(n = 100, intercept = 0, slopes = 0.5, seed = 42)

identical(data1, data2)  # TRUE
```

### 3. Input Validation

```{r validation, eval=FALSE}
# Clear error messages
simulate_garden_paths(
  observations = c(1, 2, 3),  # Invalid: not 0/1
  possibilities = c(0, 1)
)
# Error: Assertion on 'observations' failed: Must have names.

# Helpful feedback
compute_icc(
  between_group_sd = -1,  # Invalid: negative
  within_group_sd = 1
)
# Error: Assertion on 'between_group_sd' failed: Element 1 is not >= 0.
```

### 4. Comprehensive Documentation

Every function has:
- Clear description
- Parameter explanations
- Return value documentation
- Executable examples
- References to book chapters

```{r docs, eval=FALSE}
?simulate_confounding
?compute_gp_posterior
?create_dag_positions
```

## Quick Start Workflows

### Workflow 1: Explore Causal Structure

```{r workflow-1, eval=FALSE}
# 1. Draw your DAG
positions <- create_dag_positions(c("Z", "X", "Y"), layout = "custom", ...)
edges <- create_dag_edges(from = c("Z", "Z", "X"), to = c("X", "Y", "Y"), ...)

# 2. Simulate from the DAG
data <- simulate_confounding(n = 1000, effect_z_x = 1, effect_z_y = 1.5, effect_x_y = 0.5)

# 3. Test different adjustment strategies
lm(y ~ x, data = data)          # Biased
lm(y ~ x + z, data = data)      # Unbiased

# 4. Visualize
ggplot(data, aes(x = x, y = y)) +
  geom_point(aes(color = z)) +
  geom_smooth(method = "lm") +
  theme_rethinking()
```

### Workflow 2: Build and Validate GLM

```{r workflow-2, eval=FALSE}
# 1. Generate data with known parameters
data <- simulate_poisson_data(n = 300, intercept = 1.5, slopes = 0.3)

# 2. Fit model
fit <- glm(y ~ x_1, data = data, family = poisson())

# 3. Check recovery
coef(fit)["x_1"]  # Should be ≈ 0.3

# 4. Compute effect sizes
rate_ratio <- compute_rate_ratio(coef(fit)["x_1"])
```

### Workflow 3: Multilevel Analysis

```{r workflow-3, eval=FALSE}
# 1. Simulate hierarchical data
data <- simulate_varying_intercepts(n_groups = 20, n_per_group = 10)

# 2. Fit with lme4 or brms
library(lme4)
fit <- lmer(y ~ 1 + (1 | group), data = data)

# 3. Extract variance components
between_sd <- sqrt(VarCorr(fit)$group[1])
within_sd <- sigma(fit)

# 4. Analyze shrinkage
group_stats <- data %>% group_by(group) %>% summarise(mean = mean(y), n = n())
shrinkage <- compute_shrinkage(group_stats$mean, group_stats$n, within_sd, between_sd)

# 5. Compute ICC
icc <- compute_icc(between_sd, within_sd)
```

## Learning Path

### Beginner
1. `vignette("getting-started")` - Chapters 2-3 basics
2. `vignette("causal-inference")` - DAGs and confounding
3. `vignette("glm-models")` - Binomial and Poisson

### Intermediate
4. `vignette("bayesian-workflow")` - Integration with brms/tidybayes
5. Chapters 8-11 functions - MCMC and GLMs
6. Chapters 12-13 functions - Multilevel basics

### Advanced
7. `vignette("advanced-topics")` - Measurement, GPs, networks
8. Chapters 14-18 functions - Complex structures
9. Custom Stan models with cmdstanr

## Resources

### Package Documentation
- GitHub: <https://github.com/Samcil/stat_rethinking_2025>
- Issues: <https://github.com/Samcil/stat_rethinking_2025/issues>

### Related Packages
- **brms**: <https://paul-buerkner.github.io/brms/>
- **tidybayes**: <https://mjskay.github.io/tidybayes/>
- **ggdist**: <https://mjskay.github.io/ggdist/>
- **bayesplot**: <https://mc-stan.org/bayesplot/>
- **rethinking**: <https://github.com/rmcelreath/rethinking>

### Books
- McElreath, R. (2020). *Statistical Rethinking* (2nd ed.)
- Gelman et al. (2020). *Bayesian Data Analysis* (3rd ed.)
- Kruschke, J. (2014). *Doing Bayesian Data Analysis*

## Getting Help

```{r help, eval=FALSE}
# Function documentation
?simulate_confounding
?compute_gp_posterior

# Package overview
help(package = "tidyrethinking")

# All vignettes
browseVignettes("tidyrethinking")

# Report bugs
# https://github.com/Samcil/stat_rethinking_2025/issues
```
